---
title: "Analysis of Chicago traffic incident data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

Data for this project was obtained from the Chicago Data Portal. It contains 600k plus observations and has 49 features. (Making this 8gb MacBook Air work)

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(mice)
library(VIM)
library(lubridate)
library(gtsummary)
traffic <- read_csv("Traffic_Crashes_-_Crashes.csv")
```

## Data cleaning and manipulation

We chose to use the crash_type variable as the response which had Injury or No injury categories. This was changed into a binary 0 if no, 1 if yes. There is other similar creation of binary classifieres for a range of variables as well as some date formatting. 

```{r}
traffic <- traffic %>%
  mutate(CRASH_YEAR = year(as.POSIXct(substr(traffic$CRASH_DATE, 1, 10), format = "%m/%d/%Y"))) %>% 
  mutate(CRASH_TYPE_DUM = ifelse(startsWith(traffic$CRASH_TYPE, "NO"), 0, 1)) %>% 
  mutate(PERIOD = ifelse(CRASH_YEAR < 2020, "pre", "post")) %>% 
  mutate(WEATHER_CLEAR = ifelse(str_detect(WEATHER_CONDITION, "CLEAR"), 1, 0)) %>% 
  mutate(LIGHTING_DARK = ifelse(str_detect(LIGHTING_CONDITION, "DARK"), 1, 0)) %>% 
  mutate(ROAD_WET = ifelse(str_detect(ROADWAY_SURFACE_COND, "DRY|UNKNOWN"), 0, 1)) %>% 
  mutate(INTERSECTION_RELATED_I = replace_na(INTERSECTION_RELATED_I, "N"))
```


## Exploratory data analysis

Getting a sense of the data. First looking at the injury rate over time.

```{r}
traffic %>% 
  group_by(CRASH_YEAR) %>% 
  summarise(injuries_year = sum(CRASH_TYPE_DUM), total_crash = n()) %>% 
  mutate(injury_rate = injuries_year / total_crash) %>%
  filter(between(CRASH_YEAR, 2018, 2021)) %>% 
  ggplot(aes(CRASH_YEAR,injury_rate)) +
  geom_line() +
  geom_text(aes(label = round(injury_rate, 3)), position = position_stack(vjust = 1.1)) + 
  ylim(0.1, 0.4)
```

There is definitely an increase in the injury rate in 2020 and it stays elevated relative to the previous years. 


```{r}
traffic %>% 
  ggplot(aes(x = PERIOD, fill = factor(CRASH_TYPE_DUM))) +
  geom_bar(position = "fill")
```

Looking at the rates in a new category - pre and post covid. Since the change seemed to have occurred in 2020 we wanted to group the data into these categories. The post covid proportion is clearly larger. We will revisit this later and test whether these proportions are statistically different. 


```{r}
traffic %>% 
  ggplot(aes(x=factor(1), stat = "bin", fill = factor(CAUSE))) +
  geom_bar(position = "fill") + 
  facet_grid(facets =. ~ PERIOD) +
  coord_polar(theta = "y")
```

This is a breakdown of the accident causes. As you can see, a ton of data is lableled "MISC Other". Might have some opportunity to clean this further in later analysis...

## Features we want to investigate

```{r}
traffic2 <- traffic %>% 
  select(CRASH_TYPE_DUM, PERIOD, WEATHER_CLEAR, LIGHTING_DARK, ROAD_WET, INTERSECTION_RELATED_I,
         DAMAGE, STREET_DIRECTION, CAUSE, CRASH_YEAR, CRASH_MONTH, CRASH_DAY_OF_WEEK, CRASH_HOUR) %>% 
  mutate(across(CRASH_TYPE_DUM:CAUSE, factor))
```


## Individual Logistic Regressions

```{r}
logistic_table <- traffic2 %>%
  select(CRASH_TYPE_DUM, PERIOD, WEATHER_CLEAR, LIGHTING_DARK, ROAD_WET, INTERSECTION_RELATED_I,
         DAMAGE, STREET_DIRECTION, CAUSE, CRASH_YEAR, CRASH_MONTH, CRASH_DAY_OF_WEEK,
         CRASH_HOUR) %>% 
  tbl_uvregression(
    method = glm,
    y = CRASH_TYPE_DUM,
    method.args = list(family = binomial),
    exponentiate = TRUE,
    pvalue_fun = ~style_pvalue(.x, digits = 3)
  )
logistic_table
```


The odds ratio for Lighting conditions, Intersection related, damage over 1500, and DUI are especially high relative to other features. 

## 2x2 table for pre and post covid injury rates

Back to test if these are statistically different

```{r}
dattable <- traffic %>% 
  group_by(PERIOD) %>% 
  summarise(injuries_period = sum(CRASH_TYPE_DUM), total_crash = n())

table2_2 <- tibble(Group = c("Pre-pandemic", "Pre-pandemic","Post-pandemic", "Post-pandemic"),
                   Injury = c("Yes", "No", "Yes", "No"),
                   Count = c(85524, 374627 - 85524, 70435, 228494 - 70435))

addmargins(xtabs(Count ~ Group + Injury, table2_2))
prop.test(c(70435, 85524), c(228494, 374627), correct = FALSE)

oddspost <- (70435/228494) / (1 - (70435/228494))
oddspre <- (85524/374627) / (1- (85524/374627))
odds_rat <- oddspost/oddspre
RRpan <- (70435/228494) / (85524/374627)
odds_rat
```

Interesting, the proportion test indicates that these are in fact different rates. Additionally, The odds of of injury post covid are 1.51 times higher than having an injury pre covid. This should definitely be investigated further. 

## Logistic model 

Settled on this slightly reduced model for the sake of interpretation. 

```{r}
log_mod <- glm(CRASH_TYPE_DUM ~ PERIOD + WEATHER_CLEAR + LIGHTING_DARK +
                 ROAD_WET + INTERSECTION_RELATED_I + DAMAGE + CAUSE,
               family = binomial,
               data = traffic2)
summary(log_mod)

```


Getting some probabilities

```{r}
logmodcoef <- coef(log_mod)
pi1 <- exp(logmodcoef[1] + logmodcoef[4] +logmodcoef[5] +
             logmodcoef[6] + logmodcoef[8] + logmodcoef[10]) / 
  (1 + exp(logmodcoef[1] + logmodcoef[4] +logmodcoef[5] +
             logmodcoef[6] + logmodcoef[8] + logmodcoef[10]))
pi2 <- exp(logmodcoef[1] + logmodcoef[2] + logmodcoef[3]) / 
  (1 + exp(logmodcoef[1] + logmodcoef[2] + logmodcoef[3]))
pi1
pi2
```


You can see how all the different factors increase the overall probability of being injured. 

For instance, if you are driving post covid at night in the rain, and in and intersection (among other things) your probability of being injured is very high - around 0.93

While if you are not in those situations the probability is much lower - around 0.22

Although this seems pretty intuitive, it is still interesting to see how much all these factors together increase the probability of injury in the data. 


## Exploring multi category logit models

Changing the response variable to Damage which has 3 levels (<500, 501-1500, >1500)

Here is the model for incidents in an intersection. Our reference category was >1500. 

```{r, message=FALSE}
library(VGAM)
mult_mod2 <- vglm(DAMAGE ~ INTERSECTION_RELATED_I,
                 family = multinomial,
                 data = traffic2)
summary(mult_mod2)


```

Now the explenations for these are slighlty tricky, but essentially the estimated odds of having damage of <500 relative to damage of >1500 is exp(-.31521) = 0.7296 

This implies that having an incident in an intersection is conducive to increased Damage dollars. 

## Future work

The logistic model does highlight important features that increase odds which seem to match our intuitive thinking about driving. However, there is an interesting time component to this data that should be investigated further using a time series, etc. 


